Connectome inferences using partial correlation from imaging calcium signal

### 2014-06-13 article.tex
Disambiguish notation between random variables and nodes V

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $V = \left\{V_1, \ldots, V_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{V}$, and by
$V^{-i,j}$ the set of all variables except $V_i$ and $V_j$.
Two random variables $V_i$ and $V_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $V_i \perp V_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{V}_i, \mathcal{V}_j|\mathcal{Z}} = p_{\mathcal{V}_i|\mathcal{Z}}
p_{\mathcal{V}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$V_i \not\perp V_j | V^{-i,j}$ between all pair of fluorescence signal
$i$ and $j$, in other words the inferred network will be
$\hat{A}: \hat{a}_{i,j} = I(V_i \not\perp V_j | V^{-i,j}), i, j \in V$.

Since all neurons are assumed to be connected to at least one neuron,
we thus have by definition $V_i \not\perp V_j$. However,  two random
variables $A$ and $B$ might be dependent taken alone ($A \not\perp B$), but not
once we observe a third variables $C$ ($A \perp B | C$).
The conditioning over all other random variables $V^{-i,j}$ is thus necessary
to avoid predicting an edge when two neurons are not directly connected.

If the random variables $V$ follows a joint Gaussian distribution
$p_\mathcal{V} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$V_i \perp V_j | V^{-i,j}$ if the partial correlation coefficient
\[
\rho_{V_i, V_j | V^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}.
\]
is equal zero. The partial correlation gives a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{V_i, V_j | V^{-i,j}}, i, j \in V$ between all
pair of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
By contrast, we also have $V_i \perp V_j$ if the Pearson correlation
coefficient
\[
\rho_{V_i,V_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

% Speak about why averaging
The quality of the signal filtering and network inference depends
good hyper-parameter selection, which are themselves dependent on the properties
of the targeted network and measurement conditions, e.g. signal to noise ratio.
In order to improve robustness, a weighted average of partial correlation
matrices was estimated from the time series by varying the hard-threshold value
$\tau$ and the design of the low-pass filter (coefficients $(c_l^k)_{l=-3}^k$).

@@

\subsection{Data filtering and Signal distribution}

\subsection{Inference of neuronal network}

\begin{table}[H]
\centering
\caption{Inverse correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.777 & 0.767 & 0.772 & 0.774 \\
PCA                 & 0.780 & 0.770 & 0.776 & 0.777 \\
Weights             & 0.780 & 0.769 & 0.776 & 0.777 \\
Weights + PCA       & 0.783 & 0.772 & 0.779 & 0.780 \\
H-T                 & 0.893 & 0.891 & 0.891 & 0.886 \\
H-T + PCA           & 0.894 & 0.892 & 0.891 & 0.886 \\
H-T + Weights       & 0.899 & 0.897 & 0.896 & 0.891 \\
H-T + Weights + PCA & 0.900 & 0.898 & 0.896 & 0.892 \\
P-P                 & 0.925 & 0.925 & 0.924 & 0.923 \\
P-P + PCA           & 0.926 & 0.926 & 0.925 & 0.923 \\
P-P + Weights       & 0.931 & 0.930 & 0.928 & 0.927 \\
P-P + Weights + PCA & 0.932 & 0.931 & 0.930 & 0.928 \\
Averaging (PCA)     & 0.943 & 0.942 & 0.942 & 0.939 \\
Stacking            & 0.944 & 0.943 & 0.942 & 0.940 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Inverse correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.070 & 0.064 & 0.068 & 0.072\\
PCA                 & 0.076 & 0.070 & 0.075 & 0.079\\
Weights             & 0.074 & 0.067 & 0.072 & 0.076\\
Weights + PCA       & 0.080 & 0.073 & 0.079 & 0.083\\
H-T                 & 0.264 & 0.260 & 0.269 & 0.241\\
H-T + PCA           & 0.266 & 0.263 & 0.273 & 0.244\\
H-T + Weights       & 0.280 & 0.273 & 0.281 & 0.251\\
H-T + Weights + PCA & 0.284 & 0.278 & 0.285 & 0.255\\
P-P                 & 0.322 & 0.324 & 0.323 & 0.312\\
P-P + PCA           & 0.347 & 0.352 & 0.355 & 0.341\\
P-P + Weights       & 0.333 & 0.334 & 0.327 & 0.313\\
P-P + Weights + PCA & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.403 & 0.404 & 0.398 & 0.388\\
Stacking            & 0.404 & 0.405 & 0.399 & 0.389\\
\bottomrule
\end{tabular}
\end{table}

>>>>

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $X = \left\{X_1, \ldots, X_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{X}$, and by
$X^{-i,j}$ the set of all variables except $X_i$ and $X_j$.
Two random variables $X_i$ and $X_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $X_i \perp X_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{X}_i, \mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$X_i \not\perp X_j | X^{-i,j}$ between all pair of fluorescence signal
$i$ and $j$, in other words the inferred network will be
$\hat{A}: \hat{a}_{i,j} = I(X_i \not\perp X_j | X^{-i,j}), \forall i, j \in V$.

Since all neurons are assumed to be connected to at least one neuron,
we thus have by definition $X_i \not\perp X_j$. However,  two random
variables $A$ and $B$ might be dependent taken alone ($A \not\perp B$), but not
once we observe a third variables $C$ ($A \perp B | C$).
The conditioning over all other random variables $X^{-i,j}$ is thus necessary
to avoid predicting an edge when two neurons are not directly connected.

If the random variables $X$ follows a joint Gaussian distribution
$p_\mathcal{X} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$X_i \perp X_j | X^{-i,j}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}.
\]
is equal zero. The partial correlation gives a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{X_i, X_j | X^{-i,j}}, \forall i, j \in V$ between all
pairs of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
By contrast, we also have $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

% Speak about why averaging
The quality of the signal filtering and network inference depends on
good hyper-parameter selection, which are themselves dependent on the properties
of the targeted network and measurement conditions, e.g. signal to noise ratio.
In order to improve robustness, a weighted average of partial correlation
matrices was estimated from the time series by varying the hard-threshold value
$\tau$ and the design of the low-pass filter (coefficients $(c_l^k)_{l=-3}^k$).

@@

% TODO
%   - discussions about the results
%   - discussions about the metrics (Gilles text)

\begin{table}[htb]
\centering
\caption{Inverse correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.777 & 0.767 & 0.772 & 0.774 \\
PCA                 & 0.780 & 0.770 & 0.776 & 0.777 \\
Weights             & 0.780 & 0.769 & 0.776 & 0.777 \\
Weights + PCA       & 0.783 & 0.772 & 0.779 & 0.780 \\
H-T                 & 0.893 & 0.891 & 0.891 & 0.886 \\
H-T + PCA           & 0.894 & 0.892 & 0.891 & 0.886 \\
H-T + Weights       & 0.899 & 0.897 & 0.896 & 0.891 \\
H-T + Weights + PCA & 0.900 & 0.898 & 0.896 & 0.892 \\
P-P                 & 0.925 & 0.925 & 0.924 & 0.923 \\
P-P + PCA           & 0.926 & 0.926 & 0.925 & 0.923 \\
P-P + Weights       & 0.931 & 0.930 & 0.928 & 0.927 \\
P-P + Weights + PCA & 0.932 & 0.931 & 0.930 & 0.928 \\
Averaging (PCA)     & 0.943 & 0.942 & 0.942 & 0.939 \\
Stacking            & 0.944 & 0.943 & 0.942 & 0.940 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htb]
\centering
\caption{Inverse correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.070 & 0.064 & 0.068 & 0.072\\
PCA                 & 0.076 & 0.070 & 0.075 & 0.079\\
Weights             & 0.074 & 0.067 & 0.072 & 0.076\\
Weights + PCA       & 0.080 & 0.073 & 0.079 & 0.083\\
H-T                 & 0.264 & 0.260 & 0.269 & 0.241\\
H-T + PCA           & 0.266 & 0.263 & 0.273 & 0.244\\
H-T + Weights       & 0.280 & 0.273 & 0.281 & 0.251\\
H-T + Weights + PCA & 0.284 & 0.278 & 0.285 & 0.255\\
P-P                 & 0.322 & 0.324 & 0.323 & 0.312\\
P-P + PCA           & 0.347 & 0.352 & 0.355 & 0.341\\
P-P + Weights       & 0.333 & 0.334 & 0.327 & 0.313\\
P-P + Weights + PCA & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.403 & 0.404 & 0.398 & 0.388\\
Stacking            & 0.404 & 0.405 & 0.399 & 0.389\\
\bottomrule
\end{tabular}
\end{table}
